Depression and suicidality as evolved credible signals of need in social conflicts

### 2021-06-14 Paper.rmd
Added some edits to consider. Each is commented below the paragraph the change is recommended for.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

<!-- Some evolutionary theorists have questioned the bargaining model based on evidence that depression is aversive to social partners [@nettle2004; @allen2006; @nesse2006; @myers2016]. These critics overlooked the central role of *conflict* in the model.  -->

@@

<!-- Compared to suicide, however, there are fewer anthropological investigations of depression [@kleinman1985] despite frequent reports of depression or depression-like behavior in a wide range of societies [@haroz2017; @liu2020] and the potential for cross-cultural research to help understand depressive states [@haroz2017]. -->

<!-- Durkheim's pioneering study of suicide [@durkheim1897] helped launch sociology as a quantitative discipline. Sociological investigations of suicide flourished through much of the 20th century, but were supplanted by medical and public health approaches as the biomedical model of mental health took hold in the late 1970's and early 1980's [@wray2011; @syme2020]. The biomedical model tends to frame suicide as a problem with the individual, e.g., as one involving psychiatric disturbance, drug addiction, or stress sensitivity [@wray2011]. This theoretical approach has made little progress: rates of suicide, which vary widely by country, have remained relatively flat in countries for which there are long-term data, such as the US, UK, Japan, and Sweden [@owidsuicide]. -->

<!-- Evolutionary theories of depression tend to either ignore suicidality, or take it as evidence that depression is dysfunctional. Wakefield, for instance, who has devoted much of his career to arguing that most cases of depression are simply severe but normal sadness and grief, nevertheless takes suicidality to indicate pathology for some cases of depression [e.g., @wakefield2017]. @nettle2004 argues that the costs of depression, including suicidality, outweigh any benefits, and therefore adaptationist models of depression are implausible. @gilbert2006 sees suicidality as an example of a dysregulated escape defense [for reviews of evolutionary theories of depression, see @Hagen2011]. -->

<!-- As with suicide, research on depression since the 1980's has been dominated by biomedical approaches of limited success [@patten2016; @ormel2019].  -->

>>>>

<!-- have suffered genuinely severe forms of adversity and -->

@@

<!-- Compared to non-depressed individuals, those with depression report about twice as many negative events [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994] -->

@@

<!-- Of the adversity-related risk factors for depression -->

@@

<!-- Therefore, problems whose solutions require substantial investment or other changes on the part of social partners will therefore often involve social conflict even if they did not start that way [@Hagen2003]. Indeed, there is increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review].  -->

@@

<!-- Due to the likelihood that the payoffs to deception increase with greater conflicts of interests, “cheap” signals of need, such as sad expressions and crying, may often not be believed when providing support is costly. We argue next that some of the most costly and mysterious symptoms of depression -- profound loss of interest in virtually all activities and suicidality -- are *credible* and adaptive signals of need when there are conflicts of interest between victims of adversity and their social partners. -->

@@

<!-- With a cooperative species like our own, ubiquitous conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits, even among closely related individuals. (return relatives line due to broader potential audience?) -->

@@

<!-- As depression often involves a profound loss of interest in virtually all activities that can jeopardize one's productivity [@dsm5], it might therefore be an evolved bargaining strategy for relatively powerless individuals in the wake of adversity and social conflict [@Hagen2003; @SymeHagen2021; see also @Watson:2002].  -->

@@

<!-- Bargaining models assume that delaying cooperation is costly so that there is an incentive to quickly agree on a division of benefits, especially for those who highly value the fruits of the cooperative endeavor.  -->

@@

<!-- If valuations are private information, however, costly delays  -->

@@

<!-- Theoretical models of depression must account for suicidality. Suicidal ideation is one of nine diagnostic criteria for a major depressive episode (MDE) [@dsm5] and is associated with depression across cultures [@haroz2017]; depression is a major risk factor for suicidal behavior [@hawton2013]; and suicidality has been invoked to justify grief-related depression symptoms as psychopathological [e.g., @pies2014].  -->

@@

<!-- Anthropology, in contrast, has long -->

@@

@@

<!-- Among those who lack alternative options, depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, may be one's best option to elicit support despite their aversiveness to others due to their role in reliable signaling need [for similar views, see @andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956; @Rosenthal1993; @firth1936; @Firth:1961; @farberow1961]. -->
